---
title: "Forecasting Arches National Park Visitation with Weather"
subtitle: "External Predictors and Model Enhancements in R (Quarto)"
author: "AREC 346 Teaching Example"
date: "2025-11-04"
format:
  html:
    toc: true
    toc-depth: 2
    theme: cosmo
    number-sections: false
code-fold: true
code-summary: "Show code"
execute:
  echo: true
  message: false
  warning: false
---

# Overview

This report extends the baseline visitation forecasting by incorporating weather variables. We pair readable intuition with collapsed code chunks so you can skim the narrative and expand the code as needed.

What you'll see:
- Data loading with weather checks
- Feature engineering (lags, moving averages, anomalies, interactions)
- Exploratory relationships: temperature/precipitation vs visitation, correlations
- Train/test split (24-month holdout)
- Methods:
  - Linear Regression (baseline vs. with weather vs. interactions)
  - ARIMA with external regressors (xreg)
  - Random Forest with weather features
- Side-by-side model comparison, visualizations, and weather impact analysis

> Data expected at `data/arches_with_weather.csv`. If missing, run `fetch_weather_data.R` first.

## Setup

We load the required packages. Uncomment the working directory line if needed.

```{r setup, message=FALSE, warning=FALSE}
#| code-summary: "Load packages and optional working directory"
# Optional: set working directory to this project (adjust path as needed)
# setwd("~/Library/CloudStorage/OneDrive-Colostate/teaching/arec_346/other_material/forecast")

library(tidyverse)
library(lubridate)
library(forecast)
library(randomForest)
library(zoo)
library(scales)
```

## 1) Load Data

We read the pre-joined visitation + weather dataset and show basic structure. If the file is missing, we stop and instruct how to obtain it.

```{r load-data}
#| code-summary: "Read arches_with_weather.csv and inspect"
cat("\n=== LOADING DATA ===\n")

if (!file.exists("data/arches_with_weather.csv")) {
  cat("Weather data not found. Please run 'fetch_weather_data.R' first.\n")
  cat("Falling back to regular forecasting without weather variables.\n")
  stop("Run fetch_weather_data.R to get weather data")
}

full_data <- read_csv("data/arches_with_weather.csv", show_col_types = FALSE)

cat("Loaded data:", nrow(full_data), "observations\n")
cat("Date range:", as.character(min(full_data$Date)), "to",
    as.character(max(full_data$Date)), "\n")

cat("\nAvailable variables:\n")
print(names(full_data))
```

## 2) Feature Engineering

We create time indices, calendar features, lags, rolling means, weather anomalies, interactions, and seasonal indicators to help models learn richer patterns.

```{r features}
#| code-summary: "Build time, lag, rolling, anomaly, and interaction features"
cat("\n=== FEATURE ENGINEERING ===\n")

full_data <- full_data %>%
  arrange(Date) %>%
  mutate(
    # Time index
    TimeIndex = row_number(),

    # Calendar features
    MonthFactor = factor(Month),
    Quarter = quarter(Date),
    MonthName = month(Date, label = TRUE),

    # Lag features (previous year same month)
    Visits_Lag12 = lag(RecreationVisits, 12),

    # Weather lags
    Temp_Lag1 = lag(TempMean, 1),
    Precip_Lag1 = lag(TotalPrecipitation, 1),

    # Rolling visit stats
    Visits_MA3 = rollmean(RecreationVisits, k = 3, fill = NA, align = "right"),
    Visits_MA12 = rollmean(RecreationVisits, k = 12, fill = NA, align = "right"),

    # Weather anomalies (relative to overall mean)
    TempAnomaly = TempMean - mean(TempMean, na.rm = TRUE),
    PrecipAnomaly = TotalPrecipitation - mean(TotalPrecipitation, na.rm = TRUE),

    # Interaction term
    TempSunshine = TempMean * AvgSunshine,

    # Seasonal weather indicators
    IsSummer = ifelse(Month %in% 6:8, 1, 0),
    IsWinter = ifelse(Month %in% c(12, 1, 2), 1, 0),
    IsPleasantWeather = ifelse(TempMean > 50 & TempMean < 85 & TotalPrecipitation < 1, 1, 0)
  )

cat("Features created\n")
```

## 3) Exploratory Analysis

We visualize how weather relates to visitation and examine a simple correlation matrix.

```{r eda, fig.width=8, fig.height=6}
#| code-summary: "Temp vs visitation, precip vs visitation, correlations"
cat("\n=== WEATHER-VISITATION RELATIONSHIPS ===\n")

p1 <- ggplot(full_data, aes(x = TempMean, y = RecreationVisits, color = MonthName)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "loess", se = FALSE, color = "black", linewidth = 1) +
  scale_y_continuous(labels = comma) +
  scale_color_viridis_d() +
  labs(title = "Temperature vs. Visitation",
       subtitle = "Colored by month",
       x = "Average Temperature (°F)",
       y = "Recreation Visits",
       color = "Month") +
  theme_minimal()

print(p1)

p2 <- ggplot(full_data, aes(x = TotalPrecipitation, y = RecreationVisits)) +
  geom_point(alpha = 0.3, color = "steelblue") +
  geom_smooth(method = "loess", color = "red", se = TRUE) +
  scale_y_continuous(labels = comma) +
  labs(title = "Precipitation vs. Visitation",
       x = "Total Precipitation (inches)",
       y = "Recreation Visits") +
  theme_minimal()

print(p2)

weather_vars <- c("RecreationVisits", "TempMean", "TotalPrecipitation",
                  "AvgSunshine", "AvgWindSpeed", "PrecipDays")
cor_matrix <- cor(full_data %>% select(all_of(weather_vars)), use = "complete.obs")

cat("\nCorrelation Matrix:\n")
print(round(cor_matrix, 3))
```

## 4) Train–Test Split

We hold out the last 24 months for evaluation to simulate future forecasting.

```{r split}
#| code-summary: "Create 24-month holdout test set"
cat("\n=== TRAIN-TEST SPLIT ===\n")

train_size <- nrow(full_data) - 24
train_data <- full_data[1:train_size, ]
test_data  <- full_data[(train_size + 1):nrow(full_data), ]

cat("Training data:", nrow(train_data), "observations\n")
cat("Test data:", nrow(test_data), "observations\n")
```

## 5) Method 1: Linear Regression with Weather

We compare three linear models: baseline (trend + month), with weather covariates, and with interactions.

```{r lm-weather}
#| code-summary: "Fit baseline vs weather vs interaction models and score"
cat("\n=== LINEAR REGRESSION WITH WEATHER VARIABLES ===\n")

lm_baseline <- lm(RecreationVisits ~ TimeIndex + MonthFactor, data = train_data)

lm_weather <- lm(RecreationVisits ~ TimeIndex + MonthFactor +
                   TempMean + TotalPrecipitation + AvgSunshine + AvgWindSpeed,
                 data = train_data)

lm_weather_full <- lm(RecreationVisits ~ TimeIndex + MonthFactor +
                        TempMean + TotalPrecipitation + AvgSunshine +
                        AvgWindSpeed + TempSunshine + IsPleasantWeather,
                      data = train_data)

cat("\n--- Model Comparison (Training) ---\n")
cat("Baseline R²:", summary(lm_baseline)$r.squared, "\n")
cat("With Weather R²:", summary(lm_weather)$r.squared, "\n")
cat("With Weather + Interactions R²:", summary(lm_weather_full)$r.squared, "\n")

cat("\n--- Best Linear Model Summary ---\n")
print(summary(lm_weather_full))

# Predictions on test set
lm_pred_baseline <- predict(lm_baseline, newdata = test_data)
lm_pred_weather  <- predict(lm_weather,  newdata = test_data)
lm_pred_weather_full <- predict(lm_weather_full, newdata = test_data)

# Accuracy helper
calc_metrics <- function(actual, predicted, model_name) {
  mae <- mean(abs(actual - predicted), na.rm = TRUE)
  rmse <- sqrt(mean((actual - predicted)^2, na.rm = TRUE))
  mape <- mean(abs((actual - predicted) / actual), na.rm = TRUE) * 100
  cat("\n", model_name, ":\n", sep = "")
  cat("  MAE: ", format(mae, big.mark = ",", digits = 2), "\n", sep = "")
  cat("  RMSE:", format(rmse, big.mark = ",", digits = 2), "\n", sep = "")
  cat("  MAPE:", round(mape, 2), "%\n", sep = "")
  return(c(MAE = mae, RMSE = rmse, MAPE = mape))
}

cat("\n--- Test Set Performance ---\n")
metrics_lm_baseline <- calc_metrics(test_data$RecreationVisits, lm_pred_baseline,  "LM Baseline (no weather)")
metrics_lm_weather  <- calc_metrics(test_data$RecreationVisits, lm_pred_weather,   "LM with Weather")
metrics_lm_full     <- calc_metrics(test_data$RecreationVisits, lm_pred_weather_full, "LM with Weather + Interactions")
```

## 6) Method 2: ARIMA with External Regressors

We provide weather variables as external regressors (xreg) to ARIMA to capture exogenous effects on visitation.

```{r arima-xreg}
#| code-summary: "Fit ARIMA baseline and ARIMA+xreg; compare metrics"
cat("\n=== ARIMA WITH EXTERNAL REGRESSORS ===\n")

train_xreg <- train_data %>% select(TempMean, TotalPrecipitation, AvgSunshine) %>% as.matrix()
test_xreg  <- test_data  %>% select(TempMean, TotalPrecipitation, AvgSunshine) %>% as.matrix()

train_ts <- ts(train_data$RecreationVisits,
               start = c(train_data$Year[1], train_data$Month[1]),
               frequency = 12)

arima_baseline <- auto.arima(train_ts, seasonal = TRUE)
arima_baseline_fc <- forecast(arima_baseline, h = 24)

cat("\nFitting ARIMA with external regressors...\n")
arima_weather <- auto.arima(train_ts, xreg = train_xreg, seasonal = TRUE)

cat("\n--- ARIMA with Weather Summary ---\n")
print(summary(arima_weather))

arima_weather_fc <- forecast(arima_weather, xreg = test_xreg, h = 24)

metrics_arima_baseline <- calc_metrics(test_data$RecreationVisits,
                                       as.numeric(arima_baseline_fc$mean),
                                       "ARIMA Baseline")
metrics_arima_weather <- calc_metrics(test_data$RecreationVisits,
                                      as.numeric(arima_weather_fc$mean),
                                      "ARIMA with Weather")
```

## 7) Method 3: Random Forest with Weather

We include both visitation lags and weather features for a flexible, non-linear model.

```{r rf-weather, fig.width=7, fig.height=5}
#| code-summary: "Train RF with visits + weather features; importance + metrics"
cat("\n=== RANDOM FOREST WITH WEATHER VARIABLES ===\n")

train_rf <- train_data %>%
  filter(!is.na(Visits_Lag12) & !is.na(Visits_MA3) & !is.na(Visits_MA12)) %>%
  select(RecreationVisits, TimeIndex, MonthFactor, Quarter,
         Visits_Lag12, Visits_MA3, Visits_MA12,
         TempMean, TotalPrecipitation, AvgSunshine, AvgWindSpeed,
         PrecipDays, TempSunshine, IsPleasantWeather)

test_rf <- test_data %>%
  filter(!is.na(Visits_Lag12) & !is.na(Visits_MA3) & !is.na(Visits_MA12)) %>%
  select(RecreationVisits, TimeIndex, MonthFactor, Quarter,
         Visits_Lag12, Visits_MA3, Visits_MA12,
         TempMean, TotalPrecipitation, AvgSunshine, AvgWindSpeed,
         PrecipDays, TempSunshine, IsPleasantWeather)

set.seed(42)
rf_weather <- randomForest(
  RecreationVisits ~ .,
  data = train_rf,
  ntree = 500,
  mtry = 4,
  importance = TRUE
)

cat("\n--- Random Forest Model ---\n")
print(rf_weather)

cat("\n--- Variable Importance ---\n")
importance_df <- as.data.frame(importance(rf_weather)) %>%
  rownames_to_column("Variable") %>%
  arrange(desc(`%IncMSE`))
print(importance_df)

varImpPlot(rf_weather, main = "Random Forest: Variable Importance\n(Including Weather)")

rf_pred_weather <- predict(rf_weather, newdata = test_rf)

metrics_rf_weather <- calc_metrics(test_rf$RecreationVisits, rf_pred_weather,
                                   "Random Forest with Weather")
```

## 8) Model Comparison

We gather the accuracy metrics across all methods and sort by MAPE.

```{r compare}
#| code-summary: "Tabulate and sort metrics across all models"
cat("\n=== COMPREHENSIVE MODEL COMPARISON ===\n")

comparison <- data.frame(
  Model = c("LM Baseline", "LM + Weather", "LM + Weather + Interactions",
            "ARIMA Baseline", "ARIMA + Weather", "RF + Weather"),
  MAE = c(metrics_lm_baseline["MAE"], metrics_lm_weather["MAE"],
          metrics_lm_full["MAE"], metrics_arima_baseline["MAE"],
          metrics_arima_weather["MAE"], metrics_rf_weather["MAE"]),
  RMSE = c(metrics_lm_baseline["RMSE"], metrics_lm_weather["RMSE"],
           metrics_lm_full["RMSE"], metrics_arima_baseline["RMSE"],
           metrics_arima_weather["RMSE"], metrics_rf_weather["RMSE"]),
  MAPE = c(metrics_lm_baseline["MAPE"], metrics_lm_weather["MAPE"],
           metrics_lm_full["MAPE"], metrics_arima_baseline["MAPE"],
           metrics_arima_weather["MAPE"], metrics_rf_weather["MAPE"])
) %>%
  mutate(across(where(is.numeric), ~round(., 2))) %>%
  arrange(MAPE)

print(comparison)

cat("\n--- Impact of Weather Variables ---\n")
lm_improvement <- (metrics_lm_baseline["MAPE"] - metrics_lm_full["MAPE"]) / metrics_lm_baseline["MAPE"] * 100
arima_improvement <- (metrics_arima_baseline["MAPE"] - metrics_arima_weather["MAPE"]) / metrics_arima_baseline["MAPE"] * 100

cat("LM MAPE improvement with weather:", round(lm_improvement, 2), "%\n")
cat("ARIMA MAPE improvement with weather:", round(arima_improvement, 2), "%\n")
```

## 9) Visualizations: Forecast Comparisons & Weather Impact

We compare model forecasts on the test set and illustrate how weather reduces error.

```{r visuals, fig.width=9, fig.height=6}
#| code-summary: "Overlay forecasts by model and show MAPE improvement"
cat("\n=== GENERATING FORECAST VISUALIZATIONS ===\n")

forecast_comparison <- test_data %>%
  mutate(
    Actual = RecreationVisits,
    `LM Baseline` = lm_pred_baseline,
    `LM + Weather` = lm_pred_weather_full,
    `ARIMA Baseline` = as.numeric(arima_baseline_fc$mean),
    `ARIMA + Weather` = as.numeric(arima_weather_fc$mean)
  ) %>%
  select(Date, Actual, `LM Baseline`, `LM + Weather`, `ARIMA Baseline`, `ARIMA + Weather`)

# RF results (may be fewer observations due to lag NA trimming)
rf_results <- test_rf %>%
  mutate(Date = test_data$Date[match(row_number(), which(!is.na(test_data$Visits_Lag12)))],
         `RF + Weather` = rf_pred_weather) %>%
  select(Date, `RF + Weather`)

forecast_comparison <- forecast_comparison %>%
  left_join(rf_results, by = "Date")

forecast_long <- forecast_comparison %>%
  pivot_longer(cols = -Date, names_to = "Model", values_to = "Visits")

p3 <- ggplot(forecast_long, aes(x = Date, y = Visits, color = Model, linetype = Model)) +
  geom_line(linewidth = 1) +
  scale_color_manual(values = c("Actual" = "black",
                                "LM Baseline" = "gray50",
                                "LM + Weather" = "blue",
                                "ARIMA Baseline" = "pink",
                                "ARIMA + Weather" = "red",
                                "RF + Weather" = "darkgreen")) +
  scale_linetype_manual(values = c("Actual" = "solid",
                                   "LM Baseline" = "dotted",
                                   "LM + Weather" = "dashed",
                                   "ARIMA Baseline" = "dotted",
                                   "ARIMA + Weather" = "dashed",
                                   "RF + Weather" = "longdash")) +
  scale_y_continuous(labels = comma) +
  labs(title = "Forecast Comparison: With vs. Without Weather Data",
       subtitle = "Test period (24 months)",
       x = "Date",
       y = "Recreation Visits") +
  theme_minimal() +
  theme(legend.position = "bottom")

print(p3)

# Weather impact bar chart
improvement_data <- data.frame(
  Model = c("Linear Regression", "ARIMA"),
  Baseline = c(metrics_lm_baseline["MAPE"], metrics_arima_baseline["MAPE"]),
  WithWeather = c(metrics_lm_full["MAPE"], metrics_arima_weather["MAPE"])
) %>%
  pivot_longer(cols = c(Baseline, WithWeather), names_to = "Version", values_to = "MAPE")

p4 <- ggplot(improvement_data, aes(x = Model, y = MAPE, fill = Version)) +
  geom_col(position = "dodge") +
  geom_text(aes(label = round(MAPE, 2)), position = position_dodge(width = 0.9), vjust = -0.5) +
  scale_fill_manual(values = c("Baseline" = "gray70", "WithWeather" = "steelblue")) +
  labs(title = "Impact of Weather Variables on Forecast Accuracy",
       subtitle = "Lower MAPE is better",
       x = "Model Type",
       y = "MAPE (%)",
       fill = "Model Version") +
  theme_minimal()

print(p4)
```

## 10) Weather Impact Analysis

We interpret model coefficients, list the most influential weather variables, and run a simple scenario analysis.

```{r weather-impact}
#| code-summary: "Coefficients and scenario analysis"
cat("\n=== WEATHER IMPACT ANALYSIS ===\n")

weather_coefs <- coef(lm_weather_full)[c("TempMean", "TotalPrecipitation", "AvgSunshine", "AvgWindSpeed")]

cat("\nWeather Variable Effects (Linear Model):\n")
cat("  Temperature (+1°F):", round(weather_coefs["TempMean"], 0), "visits\n")
cat("  Precipitation (+1 inch):", round(weather_coefs["TotalPrecipitation"], 0), "visits\n")
cat("  Sunshine (+1 hour/day):", round(weather_coefs["AvgSunshine"], 0), "visits\n")
cat("  Wind speed (+1 mph):", round(weather_coefs["AvgWindSpeed"], 0), "visits\n")

cat("\n=== SCENARIO ANALYSIS ===\n")

baseline_scenario <- data.frame(
  TimeIndex = max(train_data$TimeIndex) + 1,
  MonthFactor = factor(7, levels = 1:12),  # July
  TempMean = 85,
  TotalPrecipitation = 0.5,
  AvgSunshine = 10,
  AvgWindSpeed = 8,
  TempSunshine = 85 * 10,
  IsPleasantWeather = 1
)

baseline_pred <- predict(lm_weather_full, newdata = baseline_scenario)

warm_scenario <- baseline_scenario
warm_scenario$TempMean <- 90
warm_scenario$TempSunshine <- 90 * 10
warm_pred <- predict(lm_weather_full, newdata = warm_scenario)

cat("Baseline (July, 85°F):", format(round(baseline_pred), big.mark = ","), "visits\n")
cat("Warmer scenario (+5°F):", format(round(warm_pred), big.mark = ","), "visits\n")
cat("Change:", format(round(warm_pred - baseline_pred), big.mark = ","),
    "(", round((warm_pred - baseline_pred) / baseline_pred * 100, 1), "%)\n\n", sep = "")

rainy_scenario <- baseline_scenario
rainy_scenario$TotalPrecipitation <- 2.0
rainy_scenario$IsPleasantWeather <- 0
rainy_pred <- predict(lm_weather_full, newdata = rainy_scenario)

cat("Rainy scenario (+1.5 inches precip):", format(round(rainy_pred), big.mark = ","), "visits\n")
cat("Change from baseline:", format(round(rainy_pred - baseline_pred), big.mark = ","),
    "(", round((rainy_pred - baseline_pred) / baseline_pred * 100, 1), "%)\n\n", sep = "")
```

## 11) Key Insights

We summarize accuracy gains and weather effects to guide decision-making.

```{r insights}
#| code-summary: "Print key takeaways"
cat("\n=== KEY INSIGHTS ===\n\n")

cat("1. WEATHER IMPACT ON FORECAST ACCURACY:\n")
cat("   - Adding weather variables improved MAPE by", round(lm_improvement, 1), "% (Linear Model)\n")
cat("   - ARIMA with weather improved by", round(arima_improvement, 1), "%\n\n")

cat("2. MOST IMPORTANT WEATHER VARIABLES:\n")
top_weather <- importance_df %>%
  filter(Variable %in% c("TempMean", "TotalPrecipitation", "AvgSunshine", "AvgWindSpeed", "PrecipDays")) %>%
  head(3)
for (i in 1:nrow(top_weather)) {
  cat("   ", i, ". ", top_weather$Variable[i], "\n", sep = "")
}
cat("\n")

cat("3. WEATHER EFFECTS ON VISITATION:\n")
if (weather_coefs["TempMean"] > 0) {
  cat("   - Warmer temperatures generally increase visitation\n")
} else {
  cat("   - Higher temperatures may decrease visitation (heat)\n")
}
if (weather_coefs["TotalPrecipitation"] < 0) {
  cat("   - Precipitation negatively impacts visitation\n")
}
if (weather_coefs["AvgSunshine"] > 0) {
  cat("   - More sunshine hours increase visitation\n")
}

cat("\n4. BEST MODEL:\n")
best_model <- comparison$Model[which.min(comparison$MAPE)]
cat("   -", best_model, "with MAPE of", min(comparison$MAPE), "%\n\n")

cat("\n=== ANALYSIS COMPLETE ===\n")
cat("Weather-enhanced forecasting models successfully trained and evaluated.\n")
```
